---
title: "Different groups in train and test data"
output: 
  pdf_document:
  toc: false
bibliography: ptLasso.bib
vignette: >
 %\VignetteIndexEntry{Different groups in train and test data}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{xgboost, survival, sparsepca, MASS}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Suppose we observe groups at test time that were unobserved at train time. For example, our training set may consist of $K$ *people* -- each with many observations -- and at test time, we wish to make predictions for observations from new people. We can still use pretraining in this setting: train a model using all data, and use this to guide the training for person-specific models.

Now however, we also fit an extra model to predict the similarity of test observations to the observations from each of the *training people*. To train this model, we use the (training) observation matrix $X$ and the response $y_{\text{sim}}$, where $y_{\text{sim}} = k$ for all observations from the $k^\text{th}$ person. When used for prediction, this model gives us a similarity (or probability) vector of length $K$ that sums to 1, describing how similar an observation is to each training person. 

At test time, we make predictions from (1) each pretrained person-specific model and (2) the person-similarity model, and we compute the weighted average of the pretrained predictions with respect to the similarity vector. Here is an example using simulated data.
```{r, echo=FALSE}
require(glmnet)
require(ptLasso)
```

```{r}
set.seed(1234)

# Start with 5 people, each with 300 observations and 200 features.
# 3 people will be used for training, and 2 for testing.
n = 300*5; p = 200;
groups = sort(rep(1:5, n/5))

# We will have different coefficients for each of the 3 training people, 
# and the first 3 features are shared support.
beta.group1 = c(-1, 1, 1, rep(0.5, 3), rep(0, p-6)); 
beta.group2 = c(-1, 1, 1, rep(0, 3), rep(0.5, 3), rep(0, p-9)); 
beta.group3 = c(-1, 1, 1, rep(0, 6), rep(0.5, 3), rep(0, p-12)); 

# The two test people are each a combination of of the training people.
# Person 4 will have observations drawn from classes 1 and 2, and
# Person 5 will have observations drawn from classes 1 and 3.
# The vector "hidden groups" is a latent variable - used to simulate data
# but unobserved in real data.
hidden.gps = groups
hidden.gps[hidden.gps == 4] = sample(c(1, 2), sum(groups == 4), replace = TRUE)
hidden.gps[hidden.gps == 5] = sample(c(1, 3), sum(groups == 5), replace = TRUE)

# We modify X according to group membership;
# we want X to cluster into groups 1, 2 and 3.
x = matrix(rnorm(n * p), nrow = n, ncol = p)
x[hidden.gps == 1, 1:3] = x[hidden.gps == 1, 1:3] + 1
x[hidden.gps == 2, 1:3] = x[hidden.gps == 2, 1:3] + 2
x[hidden.gps == 3, 1:3] = x[hidden.gps == 3, 1:3] + 3

# And now, we compute y using betas 1, 2 and 3: 
x.beta = rep(0, n)
x.beta[hidden.gps == 1] = x[hidden.gps == 1, ] %*% beta.group1 
x.beta[hidden.gps == 2] = x[hidden.gps == 2, ] %*% beta.group2 
x.beta[hidden.gps == 3] = x[hidden.gps == 3, ] %*% beta.group3
y = x.beta + 5 * rnorm(n)

```


We're ready to split into train, validation and test sets. We will use people 1, 2 and 3 for training and validation (two-thirds train, one-third validation), and people 4 and 5 for testing.
```{r}
trn.index = groups < 4
val.sample = sample(1:sum(trn.index), 1/3 * sum(trn.index), replace = FALSE)

xtrain = x[trn.index, ][-val.sample, ]
ytrain = y[trn.index][-val.sample]
gpstrain = groups[trn.index][-val.sample]

xval   = x[trn.index, ][val.sample, ] 
yval = y[trn.index][val.sample]
gpsval = groups[trn.index][val.sample]

xtest  = x[!trn.index, ]
ytest = y[!trn.index]
gpstest = groups[!trn.index]
```

We start with pretraining, where the person ID is the grouping variable.
```{r}
cvfit = cv.ptLasso(xtrain, ytrain, gpstrain, 
                   type.measure = "mse", 
                   group.intercepts = FALSE,
                   overall.lambda = "lambda.1se")
```

Now, we train a model to predict the person ID from the covariates. Because this example is simulated, we can measure the performance of our model on test data (via the confusion matrix comparing predicted group labels to true labels). In real settings, this would be impossible.
```{r}
simmod = cv.glmnet(xtrain, as.factor(gpstrain), family = "multinomial")

# Peek at performance on test data.
# Not possible with real data.
class.preds = predict(simmod, xtest, type="response")[, , 1]
table(apply(class.preds, 1, which.max), 
      hidden.gps[groups >= 4])
```


Finally we can make predictions: we have everything we need. For each test observation, we will get the pretrained prediction for all 3 training classes. Our final predictions are the weighted combination of the predictions from `ptLasso` and the class predictions from `glmnet`.
```{r}
alphahat  = cvfit$alphahat
bestmodel = cvfit$fit[[which(cvfit$alphalist == alphahat)]]
cat("Chosen alpha is", alphahat, ".\n")

offset = (1-alphahat) * predict(bestmodel$fitoverall, xtest, s = "lambda.1se")

# Get the prediction for all three classes for each test observation. 
# This will be a matrix with three columns; one for each class.
pretrained.preds = do.call(cbind, 
                        lapply(1:3, 
                               function(i) predict(bestmodel$fitpre[[i]], 
                                                   xtest,
                                                   newoffset = offset)
                      )
)

assess.glmnet( rowSums(pretrained.preds * class.preds), newy = ytest)$mse
```


There are two reasonable baselines. The first is the overall model with no grouping at all, and the second is the set of individual models (one for each group). 
```{r}
########################################################
# Baseline 1: overall model
########################################################
overall.predictions = predict(cvfit$fitoverall, xtest)
assess.glmnet(overall.predictions, newy = ytest)$mse

########################################################
# Baseline 2: individual models
########################################################
individual.preds = do.call(cbind, 
                           lapply(1:3, 
                                  function(i) predict(bestmodel$fitind[[i]], 
                                                      xtest,
                                                      type = "response")
                      )
)
assess.glmnet(rowSums(individual.preds * class.preds), newy = ytest)$mse
```

What we have done -- taking a weighted average of predictions with respect to similarity to each person -- makes sense mathematically. However, we have found better empirical results if we instead train a supervised learning algorithm to make the final prediction $\hat{y}$ using the pretrained model predictions and the class similarity predictions as features. So, let's do that here, using our so-far-untouched validation set.
```{r}
val.offset = predict(bestmodel$fitoverall, xval, s = "lambda.1se")
val.offset = (1 - alphahat) * val.offset
val.preds = do.call(cbind, 
                    lapply(1:3, function(i) predict(bestmodel$fitpre[[i]], 
                                                    xval,
                                                    newoffset = val.offset,
                                                    type = "response")
                      )
)
val.class.preds = predict(simmod, xval)[, , 1]

pred.data = cbind(val.preds, val.class.preds, val.preds * val.class.preds) 
final.model = cv.glmnet(pred.data, rowSums(val.preds * val.class.preds))

pred.data.test = cbind(pretrained.preds, 
                       class.preds, 
                       pretrained.preds * class.preds)
assess.glmnet(predict(final.model, pred.data.test), newy = ytest)$mse
```

Comparing performance of all models side-by-side shows that (1) using input groups improved performance -- including for the individual models and (2) including the final model did not help performance (but we still recommend trying this with real data).
```{r}
rd = function(x) round(x, 2)

cat("Overall model PSE: ", 
    rd(assess.glmnet(overall.predictions, newy = ytest)$mse))

cat("Individual model PSE: ", 
    rd(assess.glmnet(rowSums(individual.preds*class.preds), newy = ytest)$mse))

cat("Pretraining model PSE: ", 
    rd(assess.glmnet(rowSums(pretrained.preds*class.preds), newy = ytest)$mse))

cat("Pretraining model + final prediction model PSE: ", 
    rd(assess.glmnet(predict(final.model, 
                          cbind(pretrained.preds, 
                                class.preds, 
                                pretrained.preds * class.preds)
                      ), 
              newy = ytest)$mse))
```