---
title: "Using nonlinear bases (glmnet only)"
output: 
  pdf_document:
    toc: false
bibliography: ptLasso.bib
vignette: >
  %\VignetteIndexEntry{Using nonlinear bases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{rpart, xgboost, survival, sparsepca, MASS}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, echo=TRUE}
require(glmnet)
require(ptLasso)
```

Suppose we have a dataset with features $X$ and response $y$, where the relationship between $X$ and $y$ is a nonlinear function of the columns of $X$. Can we still use the lasso? Yes! We can *pretrain* our linear model using `xgboost` to obtain basis functions (features). Let's walk through an example.

### Example 1: xgboost pretraining 

```{r}
require(xgboost)
```

We start by simulating data ($n = 1800$, $p = 1000$) with a continuous response. Our coefficients $\beta$ are sparse; the first 200 entries will be drawn from a standard univariate normal, and the remainder are $0$. We define $y$ as $y = 1(X > 0) \beta + \epsilon$, where $\epsilon$ is noise; we hope that `xgboost` will learn the splits corresponding to $X > 0$.

```{r}
set.seed(1234)

n = 1800; p = 1000; noise = 5;

x     = matrix(rnorm(n * p), nrow=n, ncol=p)
xtest = matrix(rnorm(n * p), nrow=n, ncol=p)

x.model     = 1*(x > 0)     
xtest.model = 1*(xtest > 0) 

beta = c(rnorm(200), rep(0, p-200))

y     = x.model %*% beta + noise * rnorm(n)
ytest = xtest.model %*% beta + noise * rnorm(n)

train.folds = sample(rep(1:10, n/10))
```

Now, we run `xgboost` to get our basis functions:
```{r}
xgbfit      = xgboost(data=x, label=y, nrounds=200, max_depth=1, verbose=0)

x.boost     = predict(xgbfit, x, predleaf = TRUE) - 1
xtest.boost = predict(xgbfit, xtest, predleaf = TRUE) - 1
```

And we are ready for model fitting with `cv.glmnet`. Our two baselines are (1) a linear model that does not pretrain with `xgboost`, and (2) `xgboost`. We find that `glmnet` together with `xgboost` outperforms `glmnet` alone and `xgboost` alone.
```{r}
cvfit = cv.glmnet(x.boost, y, type.measure = "mse", foldid = train.folds)
cvfit.noboost = cv.glmnet(x, y, type.measure = "mse", foldid = train.folds)

cat("Lasso with xgboost pretraining PSE: ", 
    assess.glmnet(cvfit, newx = xtest.boost, newy = ytest)$mse)

cat("Lasso without xgboost pretraining PSE: ", 
    assess.glmnet(cvfit.noboost, newx = xtest, newy = ytest)$mse)

cat("xgboost alone PSE: ", 
    assess.glmnet(predict(xgbfit, xtest), newy = ytest)$mse)
```


### Example 2: xgboost pretraining with input groups

Now, let's repeat the above supposing our data have input groups. The only difference here is that we will use `cv.ptLasso` for our model instead of `cv.glmnet`, and we will use the group indicators as a feature when fitting `xgboost`.

We start by simulating data with 3 groups ($600$ observations in each group) and a continuous response. As before, we will simulate $y$ as $y = 1(X > 0) \beta + \epsilon$, only now we have a different $\beta$ for each group. The coefficients for the groups are in Table \@ref(tab:nonlinear). 

```{r nonlinear, echo = FALSE}
coef.table = cbind(cbind(
  rep(2, 3),
  c(1, 0, 0),
  c(0, 1, 0),
  c(0, 0, 1)
), 0)
rownames(coef.table) = paste0("group ", 1:3)
colnames(coef.table) = c("1-50", "51-100", "101-150", "151-200", "201-500")

knitr::kable(
  coef.table, booktabs = TRUE,
  caption = 'Coefficients for simulating data for use with xgboost pretraining'
)
```


```{r}
set.seed(1234)

n = 1800; p = 500; k = 3;
noise = 5;

groups = groupstest = sort(rep(1:k, n/k))

x     = matrix(rnorm(n * p), nrow=n, ncol=p)
xtest = matrix(rnorm(n * p), nrow=n, ncol=p)

x.model     = 1*(x > 0)     
xtest.model = 1*(xtest > 0)

common.beta = c(rep(2, 50), rep(0, p-50))
beta.1 = c(rep(0, 50),  rep(1, 50), rep(0, p-100)) 
beta.2 = c(rep(0, 100), rep(1, 50), rep(0, p-150)) 
beta.3 = c(rep(0, 150), rep(1, 50), rep(0, p-200)) 

y = x.model %*% common.beta + noise * rnorm(n)
y[groups == 1] = y[groups == 1] + x.model[groups == 1, ] %*% beta.1
y[groups == 2] = y[groups == 2] + x.model[groups == 2, ] %*% beta.2
y[groups == 3] = y[groups == 3] + x.model[groups == 3, ] %*% beta.3

ytest = xtest.model %*% common.beta + noise * rnorm(n)
ytest[groups == 1] = ytest[groups == 1] + xtest.model[groups == 1, ] %*% beta.1
ytest[groups == 2] = ytest[groups == 2] + xtest.model[groups == 2, ] %*% beta.2
ytest[groups == 3] = ytest[groups == 3] + xtest.model[groups == 3, ] %*% beta.3
```

Here are the dummy variables for our group indicators; we will use them to fit and predict with `xgboost`.
```{r}
group.ids     = model.matrix(~as.factor(groups) - 1) 
grouptest.ids = model.matrix(~as.factor(groupstest) - 1) 
colnames(grouptest.ids) = colnames(group.ids)
```

Now, let's train `xgboost` and `predict` to get our new features. Note that we now use `max_depth = 2`: this is intended to allow interactions between the group indicators and the other features.
```{r}
xgbfit      = xgboost(data=cbind(x, group.ids), label=y, 
                      nrounds=200, max_depth=2, verbose=0)

x.boost     = predict(xgbfit, cbind(x, group.ids), predleaf = TRUE) - 1
xtest.boost = predict(xgbfit, cbind(xtest, grouptest.ids), predleaf = TRUE) - 1
```

Finally, we are ready to fit two models trained with `cv.ptLasso`: one uses the xgboost features and the other does not. As before, we find that pretraining with xgboost improves performance relative to (1) model fitting in the original feature space and (2) xgboost alone.
```{r}
cvfit = cv.ptLasso(x.boost, y, groups=groups, type.measure = "mse")
preds = predict(cvfit, xtest.boost, groups=groupstest, alphatype = "varying")
preds = preds$yhatpre

cvfit.noboost = cv.ptLasso(x, y, groups=groups, type.measure = "mse")
preds.noboost = predict(cvfit.noboost, xtest, groups=groupstest, 
                        alphatype = "varying")
preds.noboost = preds.noboost$yhatpre

cat("ptLasso with xgboost pretraining PSE: ", 
    assess.glmnet(preds, newy = ytest)$mse)

cat("ptLasso without xgboost pretraining PSE: ", 
    assess.glmnet(preds.noboost, newy = ytest)$mse)

cat("xgboost alone PSE: ", 
    assess.glmnet(predict(xgbfit, xtest), newy = ytest)$mse)
```

